Multidimensional Persistence Module Classification via Lattice-Theoretic Convolutions

### 2020-10-10 paper/paper.tex
submitted

% to avoid loading the natbib package, add option nonatbib:
\usepackage[nonatbib]{neurips_2020_tda}

@@

Multidimensional persistence generalizes single-dimensional persistent homology
in order to tackle filtrations parameterized in multiple dimensions.
% Such
% filtrations arise, for example, in a Rips filtration of a finite metric space
% for which the finite metric space itself carries a filter function. %is parameterized by a signal
% %attributing a vector in $\R^d$ to every point of the metric space.
% Restricting
% to points in the metric space appearing below a certain coordinate-wise
% threshold results in a multi-filtration of the corresponding Rips complexes.
%Multidimensional persistence boasts stability properties similar to those in one
%dimension~\cite{lesnick_theory_2015}.
Unfortunately, there is no complete
compact barcode-like characterization of multidimensional persistence
modules~\cite{carlsson_theory_2009}. We must make do with incomplete invariants.
There are various algebraic invariants; in this paper we will use the
\textit{Hilbert function} and the \textit{multi-graded Betti numbers}, both of
which are $\mathbb{N}$-valued functions on the parameter space.
%Their names arise from their relationship with the homological algebra of
%persistence modules.
The Hilbert function is nothing more than the pointwise
(topological) Betti numbers\footnote{\textit{Caveat lector:} the multi-graded
Betti numbers are not the same as the topological Betti numbers
$\beta_i(\mathbb{X}) = \dim(H_i(\mathbb{X}))$.}, and the multi-graded Betti
numbers have a geometric interpretation in terms of births and deaths~\cite{knudson_refinement_2008}.

@@

Applying the simplicial homology functor (with coefficients in a field $k$)
$H_i$ to $\text{Rips}_r(\mathcal M)$ produces a sequence of vector spaces
$PH_i(r)$. The inclusions $\text{Rips}_r(\mathcal M) \to
\text{Rips}_{r'}(\mathcal M)$ induce maps $PH_i(r) \to PH_i(r')$, producing the
data of \introduce{persistence module}. This structure can be compactly
described as a functor from $\R$, viewed as a category via its standard order
structure, to the category $\Vect_k$ of vector spaces over $k$. The simplicity
of the category $\R$ gives these persistence modules simple structure: they
decompose as direct sums of interval modules $I_{[a,b)}$, which have
$I_{[a,b)}(r) = k$ for $a \leq r < b$ and zero otherwise. The maps are the
identity where possible and the zero map otherwise.
% is this the right shape for the intervals?

@@

\subsection{Multiparameter persistence}
The Rips construction produces a filtration of simplicial complexes from a
finite metric space; it is natural to consider the behavior of the homology
functor over a pair of coherent filtrations. Consider
a finite metric space $(\mathcal M, d)$ and a filtration
function $\rho:\mathcal M \to \R$. This data specifies a bifiltration of
simplicial complexes given by
\[\mathbb{X}_{r,t} = \text{Rips}_r\{x \in \mathcal M \mid \rho(x) \leq t\}.\]
There is a natural inclusion $\mathbb{X}_{r,t} \hookrightarrow
\mathbb{X}_{r',t'}$ whenever $(r,t) \leq (r',t')$ in the lattice $\R \times R$.
Composing with the homology functor produces a 2-parameter persistence module
\[PH_{i}: \R_+ \times \R \to \Vect; \quad (r,t) \mapsto H_i(X_{r,t}).\]

@@

While there does not exist a discrete set of invariants for $M$, we can extract
meaningful features. Two particularly informative types of features are 
 the \introduce{Hilbert function}
  \[\text{Hilb}: R_+ \times R \to \Z_+;\quad \mathbf{x} \mapsto \dim(M_\mathbf{x}),\] and
the \introduce{multi-graded Betti numbers} $\xi_j: L \to \Z_+$, for $j =
0,1,2$. For $M = PH_i$ as above, the Hilbert function counts the number of
connected components ($i=0$), cycles ($i=1$), or higher dimensional voids ($i >
1$) of the complex $\mathbb{X}_{r,t}$ at each $(r,t) \in R_+\times \R$. The multi-graded
Betti numbers, on the other hand, capture information about births
and deaths of homology classes.

@@

The data of a multidimensional persistence module is also indexed by $\R^n$ or a
regular finite subset thereof, but its natural algebraic structure is not that
of an abelian group. Rather, with its partial order structure, the indexing set
is a lattice. In processing signals associated with the persistence module, it
may be useful to take this structure into account rather than imposing the
abelian group structure implied by standard convolutions.

@@

\section{Experiments}
\begin{figure}[b]
  \caption{Comparison training curves for the lattice neural network and
    standard convolutional neural network. Results are averaged over 10 runs.}\label{fig:comparison}
    \begin{center}
    \includegraphics[width=0.49\textwidth]{lattice_accuracy.png}
    \vspace{0.2cm}
    \includegraphics[width=0.49\textwidth]{conv_accuracy.png}
    \end{center}

\end{figure}

We use a small portion of the Princeton ModelNet
dataset~\cite{zhirong_wu_3d_2015} as a source of finite
metric spaces. This dataset consists of hundreds of 3-dimensional CAD models
representing objects from 10 classes. We select two of the classes and sample
points from the 3D models to produce finite metric spaces embedded in $\R^3$.
We then compute the corresponding multidimensional persistence modules, from
which we produce features used as an input to a convolutional neural net
classifier.

The pipline thus begins with a 3D polyhedral model, of which $3000$ vertices are sampled to produce a point
cloud in $\R^3$. This point cloud then produces a bifiltered simplicial complex,
whose persistent homology we calculate using
RIVET~\cite{lesnick_interactive_2015}, sampled at a discrete grid of $40 \times
40$ points, producing lattice-indexed signals given by the Hilbert function and
the multi-graded Betti numbers $\xi_0, \xi_1, \xi_2$--four features in total. These are then passed to the classifier, which
produces a class prediction, in this case either ``sofa'' or ``monitor.''

As the filter function on these data sets, we use the \introduce{codensity} function
\[\rho_{\text{codense}}(x;k) = \frac{1}{\frac{1}{k} \sum_{y \in N_k(x)}
    d(x,y)},\]
where $N_k(x)$ is the set of the $k$ nearest neighbors to $x$.

@@

We compare the performance of two convolutional networks on this classification
task. One uses the lattice-convolution based layers described in
Section~\ref{sec:latticeCNN}, and the other uses standard convolutional layers.
Each has three convolutional layers followed by three fully connected layers.
The lattice-based convolution layers are of the form $\text{MeetConv}(x) +
\text{JoinConv}(x)$. All convolution kernels have dimension $4 \times 4$, hidden
convolution layers have 16 features, and the final convolution has 8 features.
The fully connected layers have 32 features at each inner layer.
The networks are trained using the Adam gradient algorithm with learning rate
$0.001$ for 50 epochs. We reserve 10\% of the data for testing. Results are
shown in Figure~\ref{fig:comparison}.

The lattice-based convolution and classical convolution networks perform
similarly on the training set. However, the lattice convolution network's
predictions generalize more accurately to the testing set ($\approx 7\%$ improvement in testing accuracy).
This improvement is  at the expense of more variability in performance during
training. %noisy training as well as dramatically increased training time.  
%% more discussion of what the data actually looks like

\section{Discussion}
Our proposed featurization for persistence modules is rather naive, but well
adapted to the use of lattice convolutions as a data processing method. The
lattice convolutional neural network shows promise as a method for classifying
features arising from a multiparameter persistence module. The algebraic
perspective on partially ordered sets, exemplified by lattices, may also offer
approaches to featurizing more complex invariants of persistence modules. In
particular, the incidence algebra may offer a natural way to represent the rank
invariant~\cite{carlsson_theory_2009} in a way amenable to convolution-like
operations.
We hope with these brief experiments to inspire further work on featurizing
multidimensional persistence for use in machine learning algorithms.
% anything else?

\bibliographystyle{plain}
\bibliography{latticenn}

>>>>

% to avoid loading the natbib package, add option nonatbib:
\usepackage[preprint]{neurips_2020_tda}

@@

Multidimensional persistence generalizes single-dimensional persistent homology
in order to tackle filtrations parameterized in multiple dimensions.
% Such
% filtrations arise, for example, in a Rips filtration of a finite metric space
% for which the finite metric space itself carries a filter function. %is parameterized by a signal
% %attributing a vector in $\R^d$ to every point of the metric space.
% Restricting
% to points in the metric space appearing below a certain coordinate-wise
% threshold results in a multi-filtration of the corresponding Rips complexes.
%Multidimensional persistence boasts stability properties similar to those in one
%dimension~\cite{lesnick_theory_2015}.
Unfortunately, there is no complete
compact barcode-like characterization of multidimensional persistence
modules~\cite{carlsson_theory_2009}. We must make do with incomplete invariants.
There are various algebraic invariants; in this paper we will use the
\textit{Hilbert function} and the \textit{multi-graded Betti numbers}\footnote{\textit{Caveat lector:} the multi-graded (algebraic) Betti numbers need not be confused with the topological Betti numbers, i.e.~the rank of homology.}, both of
which are $\mathbb{N}$-valued functions on the parameter space.
%Their names arise from their relationship with the homological algebra of
%persistence modules.
The Hilbert function is nothing more than the pointwise
(topological) Betti numbers, and the multi-graded Betti
numbers have a geometric interpretation in terms of births and deaths~\cite{knudson_refinement_2008}.

@@

Applying the simplicial homology functor (with coefficients in a field $k$)
$H_i$ to $\text{Rips}_r(\mathcal M)$ produces a sequence of vector spaces
$PH_i(r)$. The inclusions $\text{Rips}_r(\mathcal M) \to
\text{Rips}_{r'}(\mathcal M)$ induce maps $PH_i(r) \to PH_i(r')$, producing the
data of a \introduce{persistence module}. This structure can be compactly
described as a functor from $\R$, viewed as a category via its standard order
structure, to the category $\Vect_k$ of vector spaces over $k$. The simplicity
of the category $\R$ gives these persistence modules simple structure: they
decompose as direct sums of interval modules $I_{[a,b)}$, which have
$I_{[a,b)}(r) = k$ for $a \leq r < b$ and zero otherwise. The maps are the
identity where possible and the zero map otherwise.
% is this the right shape for the intervals?

@@

\subsection{Multiparameter persistence}
The Rips construction produces a filtration of simplicial complexes from a
finite metric space; it is natural to consider the behavior of the homology
functor over a pair of coherent filtrations. Consider
a finite metric space $(\mathcal M, d)$ and a filtration
function $\rho:\mathcal M \to \R$. This data specifies a bifiltration of
simplicial complexes given by
\[\mathbb{X}_{r,t} = \text{Rips}_r\{x \in \mathcal M \mid \rho(x) \leq t\}.\]
There is a natural inclusion $\mathbb{X}_{r,t} \hookrightarrow
\mathbb{X}_{r',t'}$ whenever $(r,t) \leq (r',t')$ in the lattice $\R \times \R$.
Composing with the homology functor produces a 2-parameter persistence module
\[PH_{i}: \R_+ \times \R \to \Vect; \quad (r,t) \mapsto H_i(X_{r,t}).\]

@@

While there does not exist a discrete set of invariants for  $PH_i$, we can extract
meaningful features. Two particularly informative types of features are 
 the \introduce{Hilbert function}
  \[\text{Hilb}: \R_+ \times \R \to \Z_+;\quad \mathbf{x} \mapsto \dim(M_\mathbf{x}),\] and
the \introduce{multi-graded Betti numbers} $\xi_j: L \to \Z_+$, for $j =
0,1,2$. For $PH_i$ as above, the Hilbert function counts the number of
connected components ($i=0$), cycles ($i=1$), or higher dimensional voids ($i >
1$) of the complex $\mathbb{X}_{r,t}$ at each $(r,t) \in \R_+\times \R$. The multi-graded
Betti numbers, on the other hand, capture information about births
and deaths of homology classes.

@@

The data of a multidimensional persistence module is also indexed by $\R^n$ or a
regular finite subset thereof, but its natural algebraic structure is not that
of an abelian group. Rather, with its partial order structure, the indexing set
is a lattice. In processing signals associated with a persistence module, it
may be useful to take this structure into account rather than imposing the
abelian group structure implied by standard convolutions.

@@

\section{Experiments}
We use a small portion of the Princeton ModelNet
dataset~\cite{zhirong_wu_3d_2015} as a source of finite
metric spaces. This dataset consists of hundreds of 3-dimensional CAD models
representing objects from 10 classes. We select two of the classes and sample
points from the 3D models to produce finite metric spaces embedded in $\R^3$.
We then compute the corresponding multidimensional persistence modules, from
which we produce features used as an input to a convolutional neural net
classifier.

The pipline thus begins with a 3D polyhedral model, of which $3000$ vertices are sampled to produce a point
cloud in $\R^3$. This point cloud then produces a bifiltered simplicial complex,
whose persistent homology we calculate using
RIVET~\cite{lesnick_interactive_2015}, sampled at a discrete grid of $40 \times
40$ points, producing lattice-indexed signals given by the Hilbert function and
the multi-graded Betti numbers $\xi_0, \xi_1, \xi_2$; four features in total. These are then passed to the classifier, which
produces a class prediction, in this case either ``sofa'' or ``monitor.''

As the filter function on these data sets, we use the \introduce{codensity} function, $\rho_{\text{codense}}(x;k) = \left( \frac{1}{k} \sum_{y \in N_k(x)} d(x,y) \right)^{-1}$,
where $N_k(x)$ is the set of the $k$ nearest neighbors to $x$; we select $k=100$.

@@

We compare the performance of two convolutional networks on this classification
task. One uses the lattice-convolution based layers described in
Section~\ref{sec:latticeCNN}, and the other uses standard convolutional layers.
Each has three convolutional layers followed by three fully connected layers.
The lattice-based convolution layers are of the form $\text{MeetConv}(x) +
\text{JoinConv}(x)$. All convolution kernels have dimension $4 \times 4$, hidden
convolution layers have 16 features, and the final convolution has 8 features.
The fully connected layers have 32 features at each inner layer.
We use a cross entropy loss function with a softmax in the final layer.
The networks are trained using the Adam gradient algorithm with learning rate
$0.001$ for 50 epochs. We reserve 10\% of the data for testing. Results are
shown in Figure~\ref{fig:comparison}.

The lattice-based convolution and classical convolution networks perform
similarly on the training set. However, the lattice convolution network's
predictions generalize more accurately to the testing set ($\approx 7\%$ improvement in testing accuracy).
This improvement is  at the expense of more variability in performance during
training. (Our code is available at \texttt{https://github.com/alfred-tarski/latticeconv} .)

\begin{figure}[h]
  \caption{Comparison training curves for the lattice neural network and
    standard convolutional neural network. Results are averaged over 5 runs.}\label{fig:comparison}
    \begin{center}
%    \includegraphics[width=0.49\textwidth]{testing-accuracy.png}
%    \vspace{0.2cm}
%    \includegraphics[width=0.49\textwidth]{training-accuracy.png}
	\includegraphics[width=0.85\textwidth]{figure.png}
    \end{center}
\end{figure}
\section{Discussion}
Our proposed featurization for persistence modules is rather naive, but well
adapted to the use of lattice convolutions as a data processing method. The
lattice convolutional neural network shows promise as a method for classifying
features arising from a multiparameter persistence module. The algebraic
perspective on partially ordered sets, exemplified by lattices, may also offer
approaches to featurizing more complex invariants of persistence modules. In
particular, the incidence algebra may offer a natural way to represent the rank
invariant~\cite{carlsson_theory_2009} in a way amenable to convolution-like
operations.
We hope with these brief experiments to inspire further work on featurizing
multidimensional persistence for use in machine learning algorithms.
% anything else?

\newpage
\small
\bibliographystyle{plain}
\bibliography{latticenn}
